\chapter{Quantum field theories as functors (KS Section 3)}

This chapter is the blueprint for Section~3 of Kontsevich--Segal,
\emph{Wick rotation and the positivity of energy in quantum field theory}
(arXiv:2105.10161), to be formalized in \texttt{KontsevichSegal/Cobordism/}
and \texttt{KontsevichSegal/FieldTheory/}. The section recasts a
$d$-dimensional field theory as a holomorphic functor on a category
$\mathcal C_d^{\mathbb C}$ of space-time manifolds carrying complex metrics.

\section{The cobordism category}

\begin{definition}[The cobordism category $\mathcal C_d^{\mathbb C}$; KS Section 3]
  \label{def:cobordism-category}
  \lean{Cobordism.CobordismGeometry, Cobordism.Semicategory,
    Cobordism.instSemicategory}
  \notready
  \uses{def:allowable}
  A \emph{complex metric} on a smooth manifold $X$ is a smooth field
  $x \mapsto g_x$ assigning to each point an allowable complex metric
  $g_x \in Q_{\mathbb C}(T_x X)$ on the tangent space. Write
  $\mathrm{Met}_{\mathbb C}(X)$ for the space of such fields.

  The category $\mathcal C_d^{\mathbb C}$ is defined as follows.

  \emph{Objects.} An object is a germ of a $d$-manifold along a closed
  two-sided $(d-1)$-manifold $\Sigma$. Concretely, $\Sigma$ is a compact
  $(d-1)$-manifold presented as a closed submanifold of a $d$-manifold $U$,
  and any two open neighbourhoods of $\Sigma$ in $U$ define the same object.
  The submanifold $\Sigma$ is two-sided in $U$ and carries a
  \emph{co-orientation}, a choice of which of its two sides is incoming and
  which is outgoing. The germ carries a complex metric, that is a germ along
  $\Sigma$ of an element $g \in \mathrm{Met}_{\mathbb C}(U)$. One usually
  suppresses the thickening $U$, the co-orientation, and the metric $g$ from
  the notation.

  \emph{Morphisms.} A morphism $\Sigma_0 \to \Sigma_1$ is a cobordism
  $M : \Sigma_0 \leadsto \Sigma_1$, a compact $d$-manifold carrying a complex
  metric $g_M \in \mathrm{Met}_{\mathbb C}(M)$, whose incoming and outgoing
  boundary germs are the objects $\Sigma_0$ and $\Sigma_1$ with their
  co-orientations and metric germs. Two cobordisms
  $M, M' : \Sigma_0 \leadsto \Sigma_1$ represent the same morphism when there
  is an isometry $M \to M'$ restricting to the identity on the germs
  $\Sigma_0$ and $\Sigma_1$.

  \emph{Composition.} Composition is concatenation of cobordisms. A cobordism
  $\Sigma_0 \leadsto \Sigma_1$ followed by a cobordism
  $\Sigma_1 \leadsto \Sigma_2$ is glued along their common boundary germ
  $\Sigma_1$ to give a cobordism $\Sigma_0 \leadsto \Sigma_2$. Taking objects
  to be germs rather than bare $(d-1)$-manifolds is exactly what makes the
  smooth structure on the concatenation well defined, since there is no
  canonical smooth structure on the concatenation of two smooth cobordisms.

  \emph{Identity morphisms.} $\mathcal C_d^{\mathbb C}$ is not literally a
  category, as it has no identity morphisms. The paper records two ways to
  deal with this, neither resolved here. One may declare an isomorphism
  $\Sigma_0 \to \Sigma_1$ to be a cobordism of zero length; such degenerate
  cobordisms exist, but in any field theory they are represented by operators
  that are not nuclear, so they fall outside the intended class of morphisms.
  The alternative, which the paper adopts, takes a field theory to be a
  functor $\Sigma \mapsto E_{\Sigma}$ from $(d-1)$-manifold germs and their
  isomorphisms to vector spaces, together with a transformation
  $Z_M : E_{\Sigma_0} \to E_{\Sigma_1}$ for each cobordism. The true
  replacement for the missing identity operators is then a condition imposed
  on field theories, namely that the canonical maps
  $\check E_{\Sigma} \to \hat E_{\Sigma}$ are injective with dense image.

  Stating this in Lean requires smooth-manifold infrastructure not developed
  here: germs of manifolds along a submanifold, two-sidedness and
  co-orientation, isometries of manifolds carrying complex metrics, and gluing
  of cobordisms. It also requires realizing $\mathrm{Met}_{\mathbb C}$ as the
  sections of a bundle whose fibre at $x$ is the domain $Q_{\mathbb C}(T_x X)$
  of allowable complex metrics from Section~2.
\end{definition}

\subsection{Analytic infrastructure}

The field-theory definition of the next section refers to several analytic
notions that Kontsevich and Segal cite as standard rather than develop: nuclear
Fréchet spaces and the trace-class maps between them, the complex manifold of
complex metrics, and holomorphic vector bundles over it. The following nodes
record the structures and properties assumed of each.

\begin{definition}[Nuclear Fréchet spaces and nuclear maps]
  \label{def:nuclear-frechet}
  \lean{FrechetSpace, IsNuclearMap, IsNuclearSpace, IsNuclearFrechetSpace,
    NuclearFrechetDuality}
  \notready
  A \emph{Fréchet space} is a complete metrizable locally convex topological
  vector space over $\mathbb C$. A continuous linear map $T \colon E \to F$
  between Fréchet spaces is \emph{nuclear} (or \emph{trace-class}) if it admits
  a representation $T x = \sum_n \lambda_n\, \phi_n(x)\, y_n$ with
  $\sum_n |\lambda_n| < \infty$, the functionals $\phi_n$ equicontinuous in the
  dual $E^*$ and the vectors $y_n$ bounded in $F$. A Fréchet space is
  \emph{nuclear} if every continuous linear map from it into a Banach space is
  nuclear.

  The properties assumed of these spaces are:
  \begin{itemize}
    \item the inverse limit of a countable sequence of nuclear maps of Fréchet
      spaces is again a nuclear Fréchet space;
    \item nuclear spaces have a unique natural tensor product (the projective
      and injective tensor products coincide), coherently associative and
      commutative, so that $E \otimes F$ is unambiguous for nuclear $E$ and $F$;
    \item the strong dual of a nuclear Fréchet space is nuclear, though usually
      not metrizable, and such spaces are reflexive, so double duality recovers
      the space;
    \item a nuclear endomorphism of such a space has a well-defined trace.
  \end{itemize}

  These are standard results in the theory of nuclear spaces, cited by
  Kontsevich and Segal and not available in Mathlib. They are assumed, not
  proved.
\end{definition}

\begin{definition}[$\mathrm{Met}_{\mathbb C}(M)$ as a complex manifold]
  \label{def:metc-complex-manifold}
  \lean{TangentStructure, MetCField, MetCManifold}
  \notready
  \uses{def:allowable, def:cobordism-category}
  For a smooth $d$-manifold $M$, the space $\mathrm{Met}_{\mathbb C}(M)$ of
  complex metrics on $M$, the smooth fields
  $x \mapsto g_x \in Q_{\mathbb C}(T_x M)$, is assumed to carry the structure of
  a complex manifold, in general infinite-dimensional. The complex structure on
  each fibre comes from $Q_{\mathbb C}(T_x M)$ being an open domain in the
  complex vector space of complex-valued quadratic forms on $T_x M$ (Section 2),
  and these fibrewise structures fit together holomorphically over the smoothly
  varying point $x$. Restriction of metrics along an inclusion of germs
  $U' \subset U$, and along the inclusion of a cobordism's boundary germ
  $\Sigma_i \hookrightarrow M$, are holomorphic maps
  $\mathrm{Met}_{\mathbb C}(U) \to \mathrm{Met}_{\mathbb C}(U')$ and
  $\mathrm{Met}_{\mathbb C}(M) \to \mathrm{Met}_{\mathbb C}(\Sigma_i)$.

  The notion of ``holomorphic'' used throughout Section 3 is the one attached to
  this complex manifold. Its complex-manifold structure, infinite-dimensional in
  general, is cited by Kontsevich and Segal and not available in Mathlib. It is
  assumed.
\end{definition}

\begin{definition}[Holomorphic vector bundles over $\mathrm{Met}_{\mathbb C}(M)$]
  \label{def:holomorphic-bundle}
  \lean{FrechetFibres, HolomorphicBundle, BundleMorphism}
  \notready
  \uses{def:metc-complex-manifold, def:nuclear-frechet}
  Over the complex manifold $\mathrm{Met}_{\mathbb C}(M)$ we assume the notion of
  a \emph{locally trivial holomorphic vector bundle} whose fibres are
  topological vector spaces, namely Fréchet spaces with nuclear comparison maps,
  together with:
  \begin{itemize}
    \item \emph{morphisms} of such bundles, fibrewise continuous linear maps
      depending holomorphically on the base point;
    \item \emph{pullback} of such a bundle along a holomorphic map of complex
      manifolds, in particular along the restriction maps
      $\mathrm{Met}_{\mathbb C}(M) \to \mathrm{Met}_{\mathbb C}(\Sigma_i)$ of the
      previous node.
  \end{itemize}

  This is the structure that the holomorphicity requirement on a field theory
  will later invoke: the spaces $E_{\Sigma}$ are to form such a bundle over
  $\mathrm{Met}_{\mathbb C}(U)$, and a cobordism map $Z_M$ is to be a morphism of
  such bundles over $\mathrm{Met}_{\mathbb C}(M)$, the bundles for the two ends
  being pulled back along the restriction maps. Locally trivial holomorphic
  bundles of topological vector spaces over an infinite-dimensional complex
  manifold are not available in Mathlib and are assumed.
\end{definition}

\section{Field theories as functors}

\begin{definition}[Field theory as a holomorphic functor; KS Section 3]
  \label{def:field-theory}
  \lean{SemicategoryFunctor, FieldTheory, NuclearFrechetFibres,
    ComplexTVSFibres}
  \notready
  \uses{def:cobordism-category, def:nuclear-frechet, def:holomorphic-bundle,
    def:metc-complex-manifold}
  A \emph{$d$-dimensional field theory} is a holomorphic functor from the
  cobordism category $\mathcal C_d^{\mathbb C}$ to the category of Fréchet
  topological vector spaces and nuclear (trace-class) linear maps, which takes
  disjoint unions to tensor products. Write $E_{\Sigma}$ for the space assigned to an object $\Sigma$ and
  $Z_M \colon E_{\Sigma_0} \to E_{\Sigma_1}$ for the nuclear (trace-class) map
  assigned to a cobordism $M \colon \Sigma_0 \leadsto \Sigma_1$. Following the
  boxed definition, $E_{\Sigma}$ is a Fréchet topological vector space. The
  paper assigns it no further structure: it plays only a provisional role,
  serving to construct the fundamental nuclear spaces between which it is
  sandwiched, and it is not asserted to be nuclear. Those fundamental nuclear
  spaces are the limit spaces $\check E_{\Sigma}$ and $\hat E_{\Sigma}$
  introduced below, whose structure differs and is recorded there.

  \emph{Induced limit structure.} Because an object $\Sigma \subset U$ is a germ
  of a $d$-manifold, it carries a direct system and an inverse system of
  neighbouring objects. Upstream, the sub-cobordisms $\Sigma' \leadsto \Sigma$
  contained in $U$, each diffeomorphic to the trivial cobordism
  $\Sigma \times [0,1]$ with outgoing end $\Sigma$, are ordered by inclusion and
  form a direct system with cobordisms to $\Sigma$. Downstream, the cobordisms
  $\Sigma \leadsto \Sigma''$ contained in $U$ form an inverse system to which
  $\Sigma$ maps. Applying the functor yields canonical maps
  \[
    \check E_{\Sigma} \;\longrightarrow\; E_{\Sigma} \;\longrightarrow\;
    \hat E_{\Sigma},
  \]
  where $\check E_{\Sigma} := \varinjlim E_{\Sigma'}$ is the direct limit over
  the upstream system and $\hat E_{\Sigma} := \varprojlim E_{\Sigma''}$ is the
  inverse limit over the downstream system. The downstream limit
  $\hat E_{\Sigma}$, being the inverse limit of a countable sequence of nuclear
  maps of Fréchet spaces, is a nuclear Fréchet space. The upstream limit
  $\check E_{\Sigma}$ is also nuclear, but in general not metrizable, so unlike
  $\hat E_{\Sigma}$ it is not a Fréchet space; the paper further identifies it as
  the strong dual of the nuclear Fréchet space $\hat E_{\Sigma^*}$, where
  $\Sigma^*$ is $\Sigma$ with its co-orientation reversed, a result deferred to a
  later node. These two limit spaces, not $E_{\Sigma}$, are the load-bearing
  nuclear spaces of the theory; $E_{\Sigma}$ exists only to construct them.

  \emph{Words requiring separate nodes.} As the paper notes, almost every word
  of this definition requires further explication. The meaning of
  \emph{holomorphic}, the \emph{continuity} property of the maps
  $\check E_{\Sigma} \to E_{\Sigma} \to \hat E_{\Sigma}$, the precise sense in
  which the functor \emph{takes disjoint unions to tensor products}, and the
  further \emph{unitarity} condition are each stated as separate nodes following
  this one. The present node fixes only the underlying functor and its induced
  limit structure.
\end{definition}

\subsection{The functor axioms}

The definition above defers four conditions to separate nodes. The first three,
holomorphicity, continuity, and the disjoint-union to tensor-product condition,
are stated here; the unitarity condition is treated separately.

\begin{definition}[Holomorphicity; KS Section 3]
  \label{def:holomorphicity}
  \lean{IsHolomorphic}
  \notready
  \uses{def:field-theory, def:holomorphic-bundle, def:metc-complex-manifold}
  The functor of a field theory is \emph{holomorphic} in the following sense.
  For a fixed smooth manifold-germ $\Sigma \subset U$, the spaces $E_{\Sigma}$
  form a locally trivial holomorphic vector bundle over the complex manifold
  $\mathrm{Met}_{\mathbb C}(U)$ of complex metrics on $U$. For a cobordism
  $M \colon \Sigma_0 \leadsto \Sigma_1$, the maps $Z_M$ form a morphism of
  holomorphic vector bundles over $\mathrm{Met}_{\mathbb C}(M)$, the bundles
  $\{E_{\Sigma_0}\}$ and $\{E_{\Sigma_1}\}$ of the two ends being pulled back to
  $\mathrm{Met}_{\mathbb C}(M)$ along the restriction maps.
\end{definition}

\begin{definition}[Continuity; KS Section 3]
  \label{def:continuity}
  \lean{IsContinuous}
  \notready
  \uses{def:field-theory}
  A field theory satisfies the \emph{continuity} property if, for every object
  $\Sigma$, the canonical maps
  \[
    \check E_{\Sigma} \;\longrightarrow\; E_{\Sigma} \;\longrightarrow\;
    \hat E_{\Sigma}
  \]
  are each injective with dense image.

  This condition is imposed on the field theory, the functor, not on the
  cobordism category. It is the true replacement for the identity morphisms that
  $\mathcal C_d^{\mathbb C}$ lacks: rather than adjoining degenerate zero-length
  cobordisms, whose operators are not nuclear, one requires the injective dense
  maps $\check E_{\Sigma} \to \hat E_{\Sigma}$ of the functor.
\end{definition}

\begin{definition}[Disjoint unions to tensor products; KS Section 3]
  \label{def:tensor-axiom}
  \lean{MonoidalCobordism, IsTensorial, partitionFunction}
  \notready
  \uses{def:field-theory, def:nuclear-frechet}
  A field theory \emph{takes disjoint unions to tensor products}: it is equipped
  with a natural isomorphism
  \[
    \check E_{\Sigma} \otimes \check E_{\Sigma'} \;\cong\;
    \check E_{\Sigma \sqcup \Sigma'},
  \]
  associative and commutative with respect to the standard isomorphisms of
  disjoint union and tensor product. The tensor product is unambiguous because
  the spaces are nuclear, nuclear spaces having a unique natural tensor product,
  and this isomorphism is equivalent to the corresponding one for the downstream
  limits, $\hat E_{\Sigma} \otimes \hat E_{\Sigma'} \cong
  \hat E_{\Sigma \sqcup \Sigma'}$. Functoriality is the assumption that
  $Z_M \otimes Z_{M'} = Z_{M \sqcup M'}$ for cobordisms $M$ and $M'$.

  This requirement forces $E_{\emptyset} = \mathbb C$ on the empty
  $(d-1)$-manifold. Hence a closed $d$-manifold $M$, regarded as a cobordism
  $\emptyset \leadsto \emptyset$, has a \emph{partition function}
  $Z_M \in \mathrm{End}(E_{\emptyset}) = \mathbb C$.
\end{definition}

\subsection{Unitarity}

The remaining condition, unitarity, is built from a dual functor, a conjugate
functor, a reality condition relating them, and a reflection-positivity
requirement. The nodes below follow that order.

\begin{definition}[Dual and conjugate functors; KS Section 3]
  \label{def:dual-conjugate-functors}
  \lean{DualConjugateGeometry, dualConj, DualRelation}
  \notready
  \uses{def:field-theory, def:cobordism-category}
  Write $\Sigma^*$ for the object $\Sigma$ with its co-orientation reversed, the
  incoming and outgoing sides exchanged while the underlying germ and metric are
  unchanged. A cobordism $M \colon \Sigma_0 \leadsto \Sigma_1$ is equally a
  cobordism $\Sigma_1^* \leadsto \Sigma_0^*$, and the two induced maps
  $E_{\Sigma_0} \to E_{\Sigma_1}$ and $E_{\Sigma_1^*} \to E_{\Sigma_0^*}$ are
  algebraic transposes of one another. Thus $\Sigma \mapsto \Sigma^*$ is a
  contravariant functor. At the level of the limit spaces,
  $\check E_{\Sigma^*}$ is the dual topological vector space of $\hat E_{\Sigma}$.

  Write $\bar\Sigma$ for the object $\Sigma$ with its complex metric $g$ replaced
  by the conjugate metric $\bar g$ and its co-orientation unchanged. Then
  $\Sigma \mapsto \bar\Sigma$ is a covariant functor. Composing the two gives the
  contravariant functor $\Sigma \mapsto \bar\Sigma^*$, which reverses the
  co-orientation and conjugates the metric.
\end{definition}

\begin{definition}[Conjugate-dual duality; KS Section 3, Appendix]
  \label{def:conjugate-dual-duality}
  \lean{PlainDuality, IsReal, ConjugateDualIdentification}
  \notready
  \uses{def:dual-conjugate-functors, def:nuclear-frechet}
  For any field theory, with no further hypothesis, the appendix to Section 3 of
  the paper proves the duality
  \[
    (\check E_{\Sigma})^* \;\cong\; \hat E_{\Sigma^*},
  \]
  that is, the strong dual of $\check E_{\Sigma}$ is $\hat E_{\Sigma^*}$. The
  companion form, that the strong dual of $\hat E_{\Sigma}$ is
  $\check E_{\Sigma^*}$, also holds. Passing between the two forms is not a mere
  swap of $\Sigma$ and $\Sigma^*$; it dualizes the first and invokes the
  reflexivity of the nuclear Fréchet space $\hat E_{\Sigma}$, which is among the
  properties assumed of nuclear spaces. We state these results; their appendix
  proof is not formalized.

  Separately, a field theory may satisfy the \emph{reality condition}: a natural
  antilinear involution
  \[
    E_{\bar\Sigma} \;\cong\; \overline{E_{\Sigma}}.
    \tag{5}
  \]
  This is an additional hypothesis on the field theory, not part of the duality
  above.

  Only for a theory satisfying (5) does the \emph{conjugate-dual} identification
  follow: the conjugate dual of $\check E_{\Sigma}$ is $\hat E_{\bar\Sigma^*}$.
  This step combines the plain duality with the reality involution (5); unlike
  the plain duality, it does not hold without (5).
\end{definition}

\begin{definition}[Unitarity; KS Section 3]
  \label{def:unitarity}
  \notready
  \uses{def:conjugate-dual-duality, def:field-theory, prop:lorentzian-boundary,
    prop:only-lorentzian}
  A germ $\Sigma$ is \emph{time-symmetric} if $\Sigma \cong \bar\Sigma^*$, that
  is, if $\Sigma \subset U$ admits a reflection with fixed-point set $\Sigma$
  that reverses the co-orientation and replaces the metric by its complex
  conjugate. The metric of a time-symmetric germ is real, and its restriction to
  the hypersurface $\Sigma$ is Riemannian.

  A field theory is \emph{unitary} if it satisfies two conditions:
  \begin{itemize}
    \item[(i)] the reality condition (5); and
    \item[(ii)] \emph{reflection-positivity}: for every time-symmetric germ
      $\Sigma \cong \bar\Sigma^*$, the hermitian duality between
      $\check E_{\Sigma}$ and $\check E_{\bar\Sigma}$ is positive-definite.
  \end{itemize}

  For a unitary theory and a time-symmetric germ $\Sigma$, the positive-definite
  hermitian form makes $\check E_{\Sigma}$ a pre-Hilbert space, which completes
  to a Hilbert space $E^{\mathrm{Hilb}}_{\Sigma}$ sitting in
  \[
    \check E_{\Sigma} \;\longrightarrow\; E^{\mathrm{Hilb}}_{\Sigma}
    \;\longrightarrow\; \hat E_{\Sigma}.
  \]

  This is where unitarity meets Section 2. The metric of a time-symmetric germ is
  real and Riemannian on the hypersurface $\Sigma$, and by
  Propositions~\ref{prop:lorentzian-boundary} and~\ref{prop:only-lorentzian} the
  nondegenerate real metrics on the Shilov boundary of $Q_{\mathbb C}(V)$ are
  exactly the Lorentzian ones. The time-symmetric hypersurfaces on which a
  unitary theory acquires Hilbert spaces are the functorial counterpart of those
  Lorentzian metrics on the Shilov boundary.
\end{definition}
