\chapter{Unitarity and global hyperbolicity (KS Section 5)}

This chapter is the blueprint for Section~5 of Kontsevich--Segal,
\emph{Wick rotation and the positivity of energy in quantum field theory}
(arXiv:2105.10161), the Lorentzian and Wick-rotation capstone, planned for
\texttt{KontsevichSegal/WickRotation/}. The Lorentzian cobordism category
$\mathcal C_d^{\mathrm{Lor}}$ lies on the boundary of the complex category
$\mathcal C_d^{\mathbb C}$, in the same sense in which the real Lorentzian metrics
lie on the Shilov boundary of $Q_{\mathbb C}(V)$. On passing to this boundary, a
unitary field theory on $\mathcal C_d^{\mathbb C}$ extends to the globally
hyperbolic Lorentzian cobordisms, where it becomes a genuine unitary theory:
time-symmetric hypersurfaces acquire Hilbert spaces, and globally hyperbolic
cobordisms between them act by unitary operators. The globally hyperbolic
cobordisms play the role that the dense open subgroup of the Shilov boundary
plays in the representation theory of a complex semigroup.

Section~4 of the paper, ``Some analogies from representation theory'', is
expository. It motivates the framework by analogy with the representation theory
of $\mathrm{PSL}_2(\mathbb R)$, of the symplectic group with its oscillator
semigroup and metaplectic representation, and of $\mathrm{Diff}^+(S^1)$, in each
of which a distinguished class of unitary representations arises as the boundary
values of holomorphic representations of a complex semigroup by contraction
operators. It states no definitions, axioms, or theorems of its own, so there is
nothing to formalize from it. Within Section~5 itself, the two-dimensional
light-line discussion, including the ``black hole'' closed light-cycle, and the
$d=1$ quantum-mechanics walkthrough are worked illustrations of the proof
technique behind Theorem~5.2; they are not separate statements to formalize.

\section{The Lorentzian cobordism category}

\begin{definition}[The Lorentzian cobordism category $\mathcal C_d^{\mathrm{Lor}}$ and $\mathcal C_d^{\mathrm{Lor},\omega}$; KS Section 5]
  \label{def:lorentzian-cobordism-category}
  \notready
  \uses{def:cobordism-category, prop:lorentzian-boundary, prop:only-lorentzian}
  The \emph{Lorentzian cobordism category} $\mathcal C_d^{\mathrm{Lor}}$ is the
  Lorentzian analogue of the complex cobordism category
  $\mathcal C_d^{\mathbb C}$.

  \emph{Objects.} An object is a \emph{Lorentzian germ}: a germ of a $d$-manifold
  along a closed co-oriented $(d-1)$-manifold $\Sigma$, carrying a Lorentzian
  metric of signature $(d-1,1)$. Exponentiating the geodesics emanating
  perpendicularly from $\Sigma$ identifies a neighbourhood with
  $\Sigma \times (-\varepsilon, \varepsilon)$, in which the metric takes the form
  $h_t - \mathrm{d}t^2$, where $t \mapsto h_t$ is a smooth family of Riemannian
  metrics on $\Sigma$.

  \emph{Morphisms.} A morphism $\Sigma_0 \to \Sigma_1$ is a \emph{Lorentzian
  cobordism} $M \colon \Sigma_0 \leadsto \Sigma_1$, a compact $d$-manifold
  carrying a Lorentzian metric whose incoming and outgoing boundary germs are the
  objects $\Sigma_0$ and $\Sigma_1$. Two cobordisms represent the same morphism
  when there is an isometry between them restricting to the identity on the
  boundary germs, exactly as for $\mathcal C_d^{\mathbb C}$.

  \emph{Composition.} Composition is concatenation of cobordisms, glued along the
  common boundary germ. As for $\mathcal C_d^{\mathbb C}$, this is not literally a
  category: there are no identity morphisms.

  \emph{The boundary relation.} $\mathcal C_d^{\mathrm{Lor}}$ lies on the boundary
  of $\mathcal C_d^{\mathbb C}$. At each point the Lorentzian metric is a real
  bilinear form of signature $(d-1,1)$, and by
  Propositions~\ref{prop:lorentzian-boundary} and~\ref{prop:only-lorentzian} the
  real Lorentzian metrics are exactly the nondegenerate real metrics on the
  Shilov boundary of $Q_{\mathbb C}(V)$. The Lorentzian germs and cobordisms are
  therefore the boundary points of the complex germs and cobordisms of
  $\mathcal C_d^{\mathbb C}$, and $\mathcal C_d^{\mathrm{Lor}}$ is the Lorentzian
  Shilov boundary of the complex cobordism category.

  \emph{The real-analytic version.} $\mathcal C_d^{\mathrm{Lor},\omega}$ is the
  subcategory in which both the manifolds and their Lorentzian metrics are
  real-analytic. This restriction is required by the operator assignment of
  Section~5: the forthcoming Principle~5.1 and Theorem~5.2 move a cobordism $M$
  inside a complexification $M_{\mathbb C}$, a holomorphic $d$-manifold
  containing $M$ as a totally real submanifold with the metric induced from a
  holomorphic symmetric form on $TM_{\mathbb C}$, and such a complexification
  exists only when $M$ and its metric are real-analytic.

  Stating this in Lean requires Lorentzian-manifold geometry not developed here:
  Lorentzian metrics of signature $(d-1,1)$ and their light-cone structure, the
  geodesic-normal form $h_t - \mathrm{d}t^2$ of a germ, isometries of Lorentzian
  cobordisms and their gluing, and, for $\mathcal C_d^{\mathrm{Lor},\omega}$, the
  real-analytic structure together with the complexification $M_{\mathbb C}$ into
  which a real-analytic cobordism embeds.
\end{definition}

\section{Global hyperbolicity}

\begin{definition}[Global hyperbolicity and the subcategory $\mathcal C_d^{\mathrm{gh}}$; KS Section 5]
  \label{def:globally-hyperbolic}
  \notready
  \uses{def:lorentzian-cobordism-category, def:allowable}
  Among the Lorentzian cobordisms of $\mathcal C_d^{\mathrm{Lor}}$ we single out the
  \emph{globally hyperbolic} ones.

  \emph{The general-relativity definition.} In general relativity a Lorentzian
  cobordism $M \colon \Sigma_0 \leadsto \Sigma_1$ between Riemannian manifolds is
  \emph{globally hyperbolic} if every maximally extended timelike geodesic in $M$
  travels from $\Sigma_0$ to $\Sigma_1$. Such an $M$ is necessarily diffeomorphic to
  $\Sigma_0 \times [0,1]$. It is exactly for globally hyperbolic $M$ that the Cauchy
  problem for the wave equation on $M$ is soluble.

  \emph{The compact definition adopted here.} Kontsevich and Segal consider only
  \emph{compact} cobordisms, which are not the usual setting of relativity theory.
  In the compact case they take global hyperbolicity to be the existence of a smooth
  \emph{time-function} $t \colon M \to [0,1]$ whose gradient, with respect to the
  Lorentzian metric, lies everywhere in the positive light-cone. Such a $t$ is then a
  fibration with Riemannian fibres, the time-slices $t^{-1}(s)$, and following the
  orthogonal trajectories to the time-slices produces a diffeomorphism
  $M \to \Sigma_0 \times [0,1]$. The two characterizations agree in the compact
  setting in which the paper works: each yields the product structure
  $M \cong \Sigma_0 \times [0,1]$, and the time-function form is the definition
  adopted.

  \emph{The globally hyperbolic subcategory.} The globally hyperbolic cobordisms are
  closed under concatenation, so the objects of $\mathcal C_d^{\mathrm{Lor}}$ together
  with the globally hyperbolic cobordisms between them form a subcategory
  $\mathcal C_d^{\mathrm{gh}} \subseteq \mathcal C_d^{\mathrm{Lor}}$. The existence of
  a time-function on a compact Lorentzian cobordism is an \emph{open} condition, so
  $\mathcal C_d^{\mathrm{gh}}$ is an \emph{open} subcategory of
  $\mathcal C_d^{\mathrm{Lor}}$. It is the analogue of the dense open subgroup of the
  Shilov boundary of a complex semigroup, the real Lie group to which the holomorphic
  contraction representations of Section~4 extend as genuine unitary operators. It
  should play that role here, although the theorem obtained below, Theorem~5.2, is
  weaker.

  \emph{The real-analytic version.} Inside the real-analytic Lorentzian category
  $\mathcal C_d^{\mathrm{Lor},\omega}$ there is the subcategory
  $\mathcal C_d^{\mathrm{gh},\omega} \subseteq \mathcal C_d^{\mathrm{Lor},\omega}$ of
  globally hyperbolic cobordisms, in which the manifolds, their Lorentzian metrics,
  and the time-function are all real-analytic. The real-analyticity of the
  time-function can be dropped, since any smooth function can be approximated
  real-analytically. This is the category required for the operator assignment of
  Section~5: the forthcoming Principle~5.1 and Theorem~5.2 move a cobordism inside a
  complexification $M_{\mathbb C}$, a holomorphic thickening that exists only when the
  cobordism and its metric are real-analytic.

  \emph{The metric normal form and the deformation into allowable metrics.} For a
  globally hyperbolic cobordism equipped with a time-function, the metric, written
  through the diffeomorphism $M \cong \Sigma_0 \times [0,1]$, takes the form
  \[ h_t + c^2 \, \mathrm{d}t^2, \qquad c \colon \Sigma_0 \times [0,1] \to \mathrm{i}\mathbb R, \]
  where $t \mapsto h_t$ is a family of Riemannian metrics on the time-slices. The
  factor $c$ is purely imaginary, recording the Lorentzian signature: $c^2$ is real
  and negative, so $h_t + c^2\,\mathrm{d}t^2 = h_t - |c|^2\,\mathrm{d}t^2$. A small
  deformation $\delta c$ pushing $c$ into the open right half-plane
  $\mathbb C_+ = \{z \in \mathbb C : \mathrm{Re}\,z > 0\}$ turns the Lorentzian metric
  into an \emph{allowable} complex metric in the sense of Section~2
  (Definition~\ref{def:allowable}): the single eigenvalue $c^2$, which for the
  Lorentzian metric sat on the negative real axis, moves off it into the complex
  plane, while the $h_t$ eigenvalues stay positive, so the angle condition holds for
  small $\delta c$. This is the deformation picture connecting
  $\mathcal C_d^{\mathrm{gh}}$ to the interior $\mathcal C_d^{\mathbb C}$. One would
  like to define the operator $Z_M$ of a Lorentzian cobordism as a limit of the
  operators attached to such deformations, but the deformed metric depends on the
  choice of time-function, which should be irrelevant to $Z_M$. Resolving this is the
  purpose of the forthcoming Principle~5.1 and Theorem~5.2.

  Stating this in Lean requires Lorentzian-manifold geometry not developed here:
  timelike geodesics and their maximal extension, the light-cone structure of a
  Lorentzian metric and the notion of the positive light-cone, smooth time-functions
  and their gradients, the condition that a gradient lie in the positive light-cone,
  fibrations with Riemannian fibres and the orthogonal trajectories giving
  $M \cong \Sigma_0 \times [0,1]$, the metric normal forms $h_t - \mathrm{d}t^2$ and
  $h_t + c^2\,\mathrm{d}t^2$, the openness of the time-function condition, and, for
  $\mathcal C_d^{\mathrm{gh},\omega}$, the real-analytic structure on the manifolds,
  metrics, and time-function together with the complexification $M_{\mathbb C}$.
\end{definition}

\section{Wick rotation of a time-symmetric germ}

\begin{definition}[Wick rotation of a time-symmetric germ; KS Section 5]
  \label{def:wick-rotation}
  \notready
  \uses{def:lorentzian-cobordism-category, def:unitarity, def:field-theory}
  For a \emph{time-symmetric} Lorentzian germ the Wick rotation continues the
  Lorentzian metric to a Riemannian one and uses it to define the Euclidean space
  $E_{\Sigma}$.

  \emph{The setup.} Let $\Sigma \subset U$ be a Lorentzian germ, with $\Sigma$
  co-oriented in $U$. Exponentiating the geodesics emanating perpendicularly from
  $\Sigma$ identifies $U$ with $\Sigma \times (-\varepsilon, \varepsilon)$, and in
  this identification the Lorentzian metric takes the geodesic-normal form
  \[ h_t - \mathrm{d}t^2, \]
  where $t \mapsto h_t$ is a smooth map from $(-\varepsilon, \varepsilon)$ into the
  manifold of Riemannian metrics on $\Sigma$, as in
  Definition~\ref{def:lorentzian-cobordism-category}.

  \emph{The Wick rotation.} Suppose the germ is \emph{time-symmetric} in the sense
  of Definition~\ref{def:unitarity}, that is $\Sigma \cong \bar\Sigma^*$. We define
  the Euclidean space $E_{\Sigma}$ of the Lorentzian germ by replacing its
  Lorentzian metric with the \emph{Wick-rotated} Riemannian metric
  \[ h_{\mathrm{i}t} + \mathrm{d}t^2, \]
  and taking $E_{\Sigma}$ to be the space that the field theory of
  Definition~\ref{def:field-theory} assigns to this Riemannian germ.

  \emph{Why time-symmetry licenses the rotation.} For a Lorentzian germ in normal
  form the time-symmetry condition $\Sigma \cong \bar\Sigma^*$ is the reflection
  $t \mapsto -t$, which reverses the co-orientation and, the metric being real,
  leaves it unchanged; invariance under the reflection is the relation
  $h_t = h_{-t}$. Thus $t \mapsto h_t$ is even, hence a function of $t^2$: there is
  a smooth $s \mapsto H(s)$ near $0$ with $h_t = H(t^2)$. The rotation
  $t \mapsto \mathrm{i}t$ replaces $t^2$ by $(\mathrm{i}t)^2 = -t^2$, so
  $h_{\mathrm{i}t} = H(-t^2)$ is defined and real, a genuine Riemannian metric on
  $\Sigma$. The rotated metric $h_{\mathrm{i}t} + \mathrm{d}t^2$ is therefore
  Riemannian, of Euclidean signature, and $E_{\Sigma}$ is well defined. Without
  time-symmetry the family $t \mapsto h_t$ need not be even, $h_{\mathrm{i}t}$ need
  not be real, and the construction does not apply.

  \emph{The limitation.} This Wick rotation defines $E_{\Sigma}$ only for
  time-symmetric germs. For a general hypersurface it does not help, and even where
  it applies the prescription is rather arbitrary, tied to the chosen normal form.
  The definition of $E_{\Sigma}$ for a general Lorentzian germ, and the independence
  of the construction from the choices made, are recovered later by the forthcoming
  Principle~5.1 and Remark~5.3.

  Stating this in Lean requires Lorentzian-manifold geometry not developed here:
  the geodesic-normal-form identification
  $U \cong \Sigma \times (-\varepsilon, \varepsilon)$ of a Lorentzian germ, the
  manifold of Riemannian metrics on $\Sigma$ together with the smooth family
  $t \mapsto h_t$, the reflection structure of a time-symmetric germ giving
  $h_t = h_{-t}$, and the analytic continuation $t \mapsto h_{\mathrm{i}t}$, that is
  the passage from $h_t = H(t^2)$ to $H(-t^2)$ which realizes the rotated metric as
  real.
\end{definition}

\section{The invariance principle (Principle 5.1)}

\begin{proposition}[Principle 5.1; KS Section 5]
  \label{prop:invariance-principle}
  \notready
  \uses{def:field-theory, def:holomorphicity, def:allowable,
    def:lorentzian-cobordism-category}
  Let $M$ be a $d$-dimensional cobordism that is a real submanifold of a complex
  $d$-manifold $M_{\mathbb C}$, and suppose the complex metric of $M$ is allowable
  and is induced from a holomorphic symmetric form $g$ on the tangent bundle
  $TM_{\mathbb C}$. Then the linear map $Z_M$ does not change when $M$ is moved
  smoothly inside $M_{\mathbb C}$ with its ends $\partial M$ held fixed, provided the
  restriction of $g$ to $M$ remains an allowable complex metric throughout the
  movement.
\end{proposition}

\begin{proof}
  \uses{def:holomorphicity, def:field-theory}
  Any infinitesimal movement of $M$ inside $M_{\mathbb C}$ is given by a complex
  vector field on $M$, that is by a complexified tangent vector to the group
  $\mathrm{Diff}(M \ \mathrm{rel}\ \partial M)$ of diffeomorphisms of $M$ fixing the
  boundary. The operator $Z_M$ is invariant under this group, which only
  reparametrizes the cobordism, so the derivative of $Z_M$ vanishes in the real
  directions of the group. By holomorphicity (Definition~\ref{def:holomorphicity})
  the operator $Z_M$ depends holomorphically on $M$, so its derivative is
  complex-linear and therefore vanishes in the complexified directions as well.
  Hence $Z_M$ does not change under the movement. This is the higher-dimensional
  form of the $d = 1$ argument, in which the operator $Z_f$ of an embedding
  $f \colon [0,1] \to \mathbb C$ depends only on the endpoint $f(1)$.
\end{proof}

Principle~5.1 is the technical engine of Section~5. To apply it one needs $M$
embedded in a complexification $M_{\mathbb C}$, which is available in the
real-analytic category $\mathcal C_d^{\mathrm{Lor},\omega}$ of
Definition~\ref{def:lorentzian-cobordism-category}. The forthcoming results use it
to define the operator $Z_M$ of a Lorentzian cobordism, to prove Theorem~5.2 that a
unitary theory induces a unitary functor on $\mathcal C_d^{\mathrm{gh},\omega}$, to
remove the dependence of the Wick rotation of Definition~\ref{def:wick-rotation} on
the choices made (Remark~5.3), and to define the spaces of observables. These are
stated as separate nodes and are not assumed here.

Stating Principle~5.1 in Lean requires geometry not developed here: the
complexification $M_{\mathbb C}$ of a real-analytic cobordism $M$ and the
realization of $M$ as a totally real submanifold of it, holomorphic symmetric forms
on the tangent bundle $TM_{\mathbb C}$ and the allowable complex metric they induce
on $M$, the smooth isotopies of $M$ inside $M_{\mathbb C}$ holding $\partial M$
fixed, and the group $\mathrm{Diff}(M \ \mathrm{rel}\ \partial M)$ with its
complexified Lie algebra of complex vector fields on $M$. The complexification
$M_{\mathbb C}$ is the heaviest assumed piece: it rests on the real-analytic
structure of $\mathcal C_d^{\mathrm{Lor},\omega}$, and the complexification of a
real-analytic manifold is not available in Mathlib.

\section{Theorem 5.2: the unitary functor on the globally hyperbolic category}

\begin{theorem}[Theorem 5.2; KS Section 5]
  \label{thm:unitary-gh}
  \notready
  \uses{def:unitarity, def:field-theory, def:continuity, def:globally-hyperbolic,
    prop:invariance-principle, def:wick-rotation}
  A unitary quantum field theory on $\mathcal C_d^{\mathbb C}$, in the sense of
  Definition~\ref{def:unitarity}, induces a functor from the real-analytic globally
  hyperbolic category $\mathcal C_d^{\mathrm{gh},\omega}$ of
  Definition~\ref{def:globally-hyperbolic} to topological vector spaces. The functor
  takes time-symmetric objects to Hilbert spaces and globally hyperbolic cobordisms
  between them to unitary operators.

  More precisely: for a time-symmetric $(d-1)$-germ $\Sigma$ there is a Hilbert space
  $E_{\Sigma}^{\mathrm{Hilb}}$ with
  \[ \check E_{\Sigma} \;\subset\; E_{\Sigma}^{\mathrm{Hilb}} \;\subset\; \hat E_{\Sigma}, \]
  the completion of the pre-Hilbert space $\check E_{\Sigma}$ of
  Definition~\ref{def:unitarity}, sitting between the limit spaces $\check E_{\Sigma}$
  and $\hat E_{\Sigma}$ of Definition~\ref{def:field-theory} through the injective
  dense maps of Definition~\ref{def:continuity}. A real-analytic globally hyperbolic
  cobordism $\Sigma_0 \leadsto \Sigma_1$ between time-symmetric hypersurfaces induces
  a unitary isomorphism
  $E_{\Sigma_0}^{\mathrm{Hilb}} \to E_{\Sigma_1}^{\mathrm{Hilb}}$ which also carries
  $\check E_{\Sigma_0}$ to $\check E_{\Sigma_1}$ and $\hat E_{\Sigma_0}$ to
  $\hat E_{\Sigma_1}$.
\end{theorem}

\begin{proof}
  \uses{prop:invariance-principle, def:globally-hyperbolic, def:wick-rotation,
    def:unitarity}
  On a time-symmetric object $\Sigma$ the reflection-positivity of
  Definition~\ref{def:unitarity} makes $\check E_{\Sigma}$ a pre-Hilbert space, whose
  completion is the Hilbert space $E_{\Sigma}^{\mathrm{Hilb}}$ of the statement.

  For a cobordism, let $M \colon \Sigma_0 \leadsto \Sigma_1$ be real-analytic and
  globally hyperbolic. Choose a time-function $t \colon M \to [0,1]$ whose level
  surfaces foliate $M$ by Riemannian hypersurfaces, and identify $M$ with
  $\Sigma_0 \times [0,1]$ by the orthogonal trajectories, as in
  Definition~\ref{def:globally-hyperbolic}. Using real-analyticity, extend $t$ and
  the metric holomorphically to a complexification $M_{\mathbb C}$, so that
  $\tau = \mathrm{i}t \colon M_{\mathbb C} \to U$ is a holomorphic fibre bundle over a
  neighbourhood $U$ of $\mathrm{i}[0,1]$, with the identification
  $\Sigma_0 \times [0,1] \to M$ extending to a holomorphic trivialization. For a
  smooth curve $f \colon [0,1] \to U$ with $f(0) = 0$ and $\mathrm{Re}\, f'(s) > 0$
  this yields a totally real submanifold $M_f$ of $M_{\mathbb C}$ lying over the
  curve and carrying an allowable complex metric, the higher-dimensional form of the
  Wick rotation of Definition~\ref{def:wick-rotation}. Using the morphism of $M_f$ as
  the operator $Z_f$ was used in the $d = 1$ contour argument gives a unitary
  operator $Z_M$ for the Lorentzian cobordism, its unitarity coming from the
  reflection-positivity of Definition~\ref{def:unitarity} as in that case.

  The operator $Z_M$ is independent of the choices. By Principle~5.1
  (Proposition~\ref{prop:invariance-principle}) it does not depend on the curve $f$,
  two curves with the same endpoints being joined by a movement of $M_f$ inside
  $M_{\mathbb C}$; and it does not depend on the time-function, since two
  time-functions are linearly homotopic and passing between them deforms the totally
  real embedding $\Sigma_0 \times [0,1] \to M_{\mathbb C}$ by a real-analytic
  diffeomorphism of $\Sigma_0 \times [0,1]$, again covered by Principle~5.1.
\end{proof}

Stating Theorem~5.2 in Lean rests on infrastructure deferred elsewhere. The source
category $\mathcal C_d^{\mathrm{gh},\omega}$ carries the real-analytic and
complexification geometry assumed in
Definitions~\ref{def:lorentzian-cobordism-category} and~\ref{def:globally-hyperbolic}
and Proposition~\ref{prop:invariance-principle}: the complexification $M_{\mathbb C}$,
the totally real submanifolds $M_f$, and the holomorphic trivialization, with
$M_{\mathbb C}$ the heaviest assumed piece. The Hilbert space
$E_{\Sigma}^{\mathrm{Hilb}}$ and the inclusions
$\check E_{\Sigma} \subset E_{\Sigma}^{\mathrm{Hilb}} \subset \hat E_{\Sigma}$ inherit
the completion that Section~3 deferred at Definition~\ref{def:unitarity}: the
pre-Hilbert form on $\check E_{\Sigma}$ is statable, but its completion and the maps
into $\hat E_{\Sigma}$ need the limit relation between $\check E_{\Sigma}$,
$E_{\Sigma}$, and $\hat E_{\Sigma}$ that Section~3 left deferred. The unitary operator
$Z_M$ rests on Principle~5.1 and hence on the same complexification.

\section{Remark 5.3: well-definedness of the Lorentzian space}

Principle~5.1 resolves the arbitrariness left open by the Wick rotation of
Definition~\ref{def:wick-rotation}: the Lorentzian space $\hat E_{\Sigma}$ does not
depend on the path used to define it. Suppose the Lorentzian metric on $U$ is
real-analytic. The complex theory then gives a holomorphic bundle $\{\hat E_f\}$ over
the space $\mathcal J$ of germs of embeddings
$f \colon (-\varepsilon, \varepsilon) \to \mathbb C$ with $f(0) = 0$ and
$\mathrm{Re}\, f'(t) > 0$ for all $t$. For $\lambda \in \mathbb C_+$ the radial path
$f_{\lambda}(t) = \lambda t$ lies in $\mathcal J$, and $\hat E_f$ is the inverse limit
of the spaces attached to the germs of $f$ at the points $f(t_k)$, for any sequence
$t_k \downarrow 0$.

\begin{proposition}[Remark 5.3; KS Section 5]
  \label{prop:lorentzian-E-welldefined}
  \notready
  \uses{prop:invariance-principle, def:wick-rotation, def:field-theory, def:unitarity,
    def:lorentzian-cobordism-category}
  Any smooth path $f \in \mathcal J$ defines the same space $\hat E_{\Sigma}$: for any
  two paths there is a canonical isomorphism between the corresponding inverse limits,
  and these isomorphisms are coherent. The family is compatible with conjugation:
  $\hat E_{\bar f}$ is the complex-conjugate space to $\hat E_f$, so reversing the
  complex time-direction conjugates the identification of $\hat E_{\Sigma}$ with the
  Euclidean choice $\hat E_{f_1}$. If the germ $\Sigma \subset U$ is time-symmetric,
  and only then, this yields a hermitian inner product on $\check E_{\Sigma}$, the
  reflection-positivity pairing of Definition~\ref{def:unitarity}.
\end{proposition}

\begin{proof}
  \uses{prop:invariance-principle, def:unitarity}
  For path-independence it suffices to identify $\hat E_{f_{\lambda}}$ with
  $\hat E_{f_{\lambda'}}$ for two neighbouring radial rays with
  $|\lambda| = |\lambda'|$. Choose a sequence $t'_k \downarrow 0$ interleaving $t_k$,
  with $t_k > t'_k > t_{k+1}$, and a path $f \in \mathcal J$ in the sector between
  $f_{\lambda}$ and $f_{\lambda'}$ coinciding with them alternately near the points
  $\lambda t_k$ and $\lambda' t'_k$. This $f$ provides cobordisms from the germ at
  $\lambda' t'_k$ to the germ at $\lambda t_k$, and from the germ at $\lambda t_{k+1}$
  to the germ at $\lambda' t'_k$; composed along the two interleaved cofinal sequences
  they give mutually inverse canonical isomorphisms between the inverse limits
  $\hat E_{f_{\lambda}}$ and $\hat E_{f_{\lambda'}}$. Their coherence for three nearby
  rays follows from Principle~5.1 (Proposition~\ref{prop:invariance-principle}), the
  two routes between rays differing by a movement of the connecting path inside the
  complexification. The conjugate-space property holds because $\bar f$ corresponds to
  the complex-conjugate metric; combined with the duality of
  Definition~\ref{def:unitarity} it gives the hermitian inner product on
  $\check E_{\Sigma}$ when $\Sigma$ is time-symmetric.
\end{proof}

Stating Remark~5.3 in Lean rests on infrastructure not developed here: the space
$\mathcal J$ of embedding-germs $f \colon (-\varepsilon, \varepsilon) \to \mathbb C$
with $f(0) = 0$ and $\mathrm{Re}\, f'(t) > 0$, the holomorphic bundle $\{\hat E_f\}$
over it, the interleaving-sequence construction of the connecting cobordisms, and the
complex-conjugate-space functor $\hat E_f \mapsto \hat E_{\bar f}$. The inverse-limit
spaces $\hat E_f$ and $\check E_{\Sigma}$ carry the topological-vector-space limit
structure that Section~3 deferred (Definition~\ref{def:field-theory}), and the
coherence of the identifications rests on Principle~5.1 and hence on the
complexification $M_{\mathbb C}$.

\section{Field operators and spacelike commutativity}

The functorial picture yields local field operators at the points of a space-time, and
recovers the Wightman axiom that operators at space-like separated points commute.

\begin{definition}[Observables at a point; KS Section 5]
  \label{def:observables}
  \notready
  \uses{def:metc-complex-manifold, prop:real-subspaces, prop:invariance-principle,
    def:wick-rotation}
  In Section~3 the space $\mathcal O_x$ of \emph{observables} at a point $x$ of a
  manifold carrying a complex metric was introduced as the inverse limit of the spaces
  $E_{\partial D}$ attached to the boundaries of the closed discs $D$ around $x$,
  equivalently the fibre at the germ of the metric of a bundle on the space
  $\mathrm{Met}_{\mathbb C}(\hat x)$ of germs of complex metrics at $x$. For a
  real-analytic Lorentzian manifold $M$ and $x \in M$, the space $\mathcal O_x$ is
  defined by Wick rotation, as in Definition~\ref{def:wick-rotation}. Embed $M$ in a
  complexification $M_{\mathbb C}$, with the holomorphic exponential map from a
  neighbourhood of $0$ in the complexified tangent space
  $T_x^{\mathbb C} = T_x M \otimes \mathbb C$ to $M_{\mathbb C}$. The $d$-dimensional
  real subspaces $V \subset T_x^{\mathbb C}$ on which the metric induced by the complex
  bilinear form is allowable form a contractible open subset
  $\mathcal U \subset \mathrm{Gr}_d(T_x^{\mathbb C})$ of the real Grassmannian, by
  Proposition~\ref{prop:real-subspaces}. Exponentiating each $V$ gives a germ of a
  $d$-manifold with a complex metric, hence a map
  $\mathcal U \to \mathrm{Met}_{\mathbb C}(\hat x)$; the pullback of the bundle of
  observables along it is a bundle on $\mathcal U$, trivial by Principle~5.1
  (Proposition~\ref{prop:invariance-principle}) as in Remark~5.3, since $\mathcal U$ is
  contractible. Identifying its fibres defines $\mathcal O_x$ for the Lorentzian $M$.
\end{definition}

\begin{proposition}[Action of observables; KS Section 5]
  \label{prop:observable-action}
  \notready
  \uses{def:observables, def:field-theory, def:tensor-axiom}
  Let $M \colon \Sigma_0 \leadsto \Sigma_1$ be a Lorentzian cobordism and
  $x \in \mathring M$. An observable $\psi \in \mathcal O_x$ acts as an operator
  $E_{\Sigma_0} \to E_{\Sigma_1}$. If $x$ lies on a time-slice $\Sigma$, then $\psi$
  acts as an element of $\mathrm{Hom}(\check E_{\Sigma}; \hat E_{\Sigma})$, an unbounded
  operator in $E_{\Sigma}$. For distinct points $x_1, \ldots, x_k$ on $\Sigma$ there is
  a multilinear map
  \[ \mathcal O_{x_1} \otimes \cdots \otimes \mathcal O_{x_k} \;\longrightarrow\;
     \mathrm{Hom}(\check E_{\Sigma}; \hat E_{\Sigma}) \]
  that does not depend on an ordering of the points.
\end{proposition}

\begin{proof}
  \uses{def:field-theory, def:tensor-axiom}
  Removing a small disc $D$ around $x$ from $M$ leaves a cobordism
  $M \setminus \mathring D \colon \partial D \sqcup \Sigma_0 \leadsto \Sigma_1$, which by
  the tensoring property (Definition~\ref{def:tensor-axiom}) gives a map
  $E_{\partial D} \to \mathrm{Hom}(E_{\Sigma_0}; E_{\Sigma_1})$; passing to the inverse
  limit over $D$ gives the action of $\mathcal O_x$ as $E_{\Sigma_0} \to E_{\Sigma_1}$.
  When $x$ lies on a time-slice $\Sigma$, the cobordisms cut out by successively thinner
  collars of $\Sigma$ give instead an element of
  $\mathrm{Hom}(\check E_{\Sigma}; \hat E_{\Sigma})$. The same construction applied to
  disjoint collars of the distinct points $x_1, \ldots, x_k$, with the tensoring
  property, yields the multilinear map; it is independent of the ordering because
  disjoint collars may be taken in any order.
\end{proof}

\begin{proposition}[Spacelike commutativity; KS Section 5]
  \label{prop:spacelike-commutativity}
  \notready
  \uses{prop:observable-action, thm:unitary-gh, def:globally-hyperbolic,
    def:field-theory}
  On a real-analytic globally hyperbolic space-time the field operators of observables
  at space-like separated points commute, which is the Wightman locality axiom.
  Precisely: by the unitary propagation $Z_{t,t'}$ of Theorem~\ref{thm:unitary-gh}, the
  rigged Hilbert spaces $\check E_{\Sigma_t} \subset E_{\Sigma_t}^{\mathrm{Hilb}} \subset
  \hat E_{\Sigma_t}$ for $t_0 \le t \le t_1$ are identified into a single rigged Hilbert
  space $\check E \subset E^{\mathrm{Hilb}} \subset \hat E$, and an observable
  $\psi \in \mathcal O_x$ with $x \in \Sigma_t$ defines an unbounded operator
  \[ \tilde\psi = Z_{t_0,t}^{-1} \circ \psi \circ Z_{t_0,t} \colon \check E \to \hat E, \]
  independent of the choice of time-function. If $x$ and $x'$ are space-like separated,
  the operators $\tilde\psi$ and $\tilde\psi'$ are composable and commute.
\end{proposition}

\begin{proof}
  \uses{thm:unitary-gh, prop:observable-action, prop:invariance-principle}
  The identification of the rigged Hilbert spaces along the foliation and the operator
  $\tilde\psi$ use the unitary propagation of Theorem~\ref{thm:unitary-gh}. That
  $\tilde\psi$ does not depend on the time-function follows from Principle~5.1
  (Proposition~\ref{prop:invariance-principle}): changing the time-function moves the
  time-slice through the complexification, leaving the operators unchanged. If $x$ and
  $x'$ are space-like separated, a single time-slice $\Sigma_t$ contains both, and the
  multilinear map of Proposition~\ref{prop:observable-action} writes the composite in
  either order as $Z_{t_0,t}^{-1} \circ (\psi \otimes \psi') \circ Z_{t_0,t}$, which is
  symmetric in $\psi$ and $\psi'$; hence $\tilde\psi$ and $\tilde\psi'$ commute.
\end{proof}

Stating these results in Lean rests on infrastructure not developed here. The space
$\mathcal O_x$ is a fibre of a bundle on $\mathrm{Met}_{\mathbb C}(\hat x)$, inheriting
the deferred complex-manifold structure of $\mathrm{Met}_{\mathbb C}$
(Definition~\ref{def:metc-complex-manifold}); its Lorentzian definition uses the
complexified tangent space $T_x^{\mathbb C}$, the real Grassmannian
$\mathrm{Gr}_d(T_x^{\mathbb C})$ and the contractible open set $\mathcal U$ of allowable
subspaces (Proposition~\ref{prop:real-subspaces}), and the complexification
$M_{\mathbb C}$ of Principle~5.1. The elements of
$\mathrm{Hom}(\check E_{\Sigma}; \hat E_{\Sigma})$ are unbounded operators in
$E_{\Sigma}$, and the rigged Hilbert space
$\check E \subset E^{\mathrm{Hilb}} \subset \hat E$ inherits the Hilbert completion and
the topological-vector-space limit structure left deferred in Section~3 and Theorem~5.2.
Space-like separation is Lorentzian causal geometry, also not in Mathlib.

\section{The vacuum-expectation domain and Proposition 5.4}

The final question of the paper concerns the analytic continuation of vacuum
expectation values in the traditional treatment over Minkowski space
$\mathbb M = \mathbb R^{d-1,1}$. The Wightman axioms make the $k$-point functions the
boundary values of holomorphic functions on the \emph{permuted extended tube}
$\mathcal U_k \subset (\mathbb M_{\mathbb C})^k$, the Wightman domain defined in
Section~2; for example $\mathcal U_2$ is the set of pairs $(x,y)$ for which
$\|x-y\|^2$ is not a real number $\le 0$. For $k > 2$ the domain $\mathcal U_k$ is
known not to be holomorphically convex, so it is not the largest manifold to which the
expectation values continue, and describing that largest manifold $\mathcal V_k$, or
even the holomorphic envelope of $\mathcal U_k$, is an old open problem.

\begin{definition}[The vacuum-expectation domain $\mathcal V_k$; KS Section 5]
  \label{def:vacuum-domain}
  \notready
  \uses{def:allowable, prop:invariance-principle, def:observables}
  The framework of this paper suggests a candidate for $\mathcal V_k$. It sits over the
  open subset $\check{\mathcal V}_k$ of those $k$-tuples
  $\mathbf x = \{x_1, \ldots, x_k\}$ of distinct points of $\mathbb M_{\mathbb C}$ that
  lie on some totally real submanifold $M$ with
  \begin{itemize}
    \item[(i)] the metric induced on $M$ from $\mathbb M_{\mathbb C}$ allowable, and
    \item[(ii)] $M$ projecting surjectively onto the real Euclidean subspace
      $\mathbb E = \mathbb R^d$ of
      $\mathbb M_{\mathbb C} = \mathbb E \oplus \mathrm i \mathbb E$.
  \end{itemize}
  When the metric of $M$ is allowable the projection $M \to \mathbb E$ is a local
  diffeomorphism, so (ii) makes $M$ the graph of a smooth map
  $\mathbb E \to \mathrm i \mathbb E$. Let $\mathcal F_k$ be the space of pairs
  $(M, \mathbf x)$ satisfying (i) and (ii), an infinite-dimensional complex manifold
  with an open projection
  $\pi \colon \mathcal F_k \to \check{\mathcal V}_k \subset
  \mathrm{Conf}_k(\mathbb M_{\mathbb C})$. Then $\mathcal V_k$ is the largest Hausdorff
  quotient manifold of $\mathcal F_k$ through which $\pi$ factorizes and which maps to
  $\check{\mathcal V}_k$ by a local diffeomorphism: two pairs $(M, \mathbf x)$ and
  $(M', \mathbf x)$ in a fibre have the same image when they lie in one connected
  component of the fibre, and more generally when they are joined to a common third
  pair by paths covering the same path in $\check{\mathcal V}_k$, the further
  identifications making the quotient Hausdorff.

  A field theory holomorphic on the enlarged cobordism category of allowable
  submanifolds of $\mathbb M_{\mathbb C}$ cut between time-symmetric hypersurfaces, with
  a Lorentz-invariant vacuum, has vacuum expectation values forming a holomorphic
  section of a bundle on $\mathcal F_k$; trivializing by the local diffeomorphism
  $M \to \mathbb E$ gives a holomorphic function
  \[ \mathcal E_k \colon \mathcal F_k \;\longrightarrow\;
     \mathrm{Hom}(\mathcal O^{\otimes k}; \mathbb C), \]
  where $\mathcal O$ is the space of observables at a point of $\mathbb E$. By
  Principle~5.1 (Proposition~\ref{prop:invariance-principle}) the value of
  $\mathcal E_k$ is unchanged when the marked points $\mathbf x$ are held fixed and $M$
  is moved smoothly within the allowable class, so $\mathcal E_k$ is constant on the
  connected components of the fibres of $\pi$, the isotopy classes of allowable
  manifolds through $\mathbf x$.
\end{definition}

Whether $\mathcal V_k$ is a domain of holomorphy is left open: Kontsevich and Segal
state that they have no proof of it. This is a conjecture, not a theorem of the paper.
What is proved is the following inclusion.

\begin{proposition}[Proposition 5.4; KS Section 5]
  \label{prop:Vk-contains-Uk}
  \notready
  \uses{def:vacuum-domain, def:allowable}
  The domain $\mathcal V_k$ contains the Wightman domain $\mathcal U_k$.
\end{proposition}

\begin{proof}
  \uses{def:allowable}
  The domain $\mathcal V_k$ is invariant under the complex orthogonal group of
  $\mathbb M_{\mathbb C}$ and under reorderings of the points $\mathbf x$, so it
  suffices to treat $\mathbf x$ with $\mathrm{Im}(x_{i+1} - x_i)$ in the forward
  light-cone $C \subset \mathbb M$ for each $i$. Smoothing the polygonal path through
  the points, assume the $x_i$ lie on a curve
  $x \colon \mathbb R \to \mathbb M_{\mathbb C}$ with $\mathrm{Im}(x'(t)) \in C$ for all
  $t$. Choose, smoothly in $t$, orthonormal vectors
  $e_1(t), \ldots, e_{d-1}(t)$ in $\mathbb M_{\mathbb C}$ orthogonal to $x'(t)$, and let
  $M_t$ be the real subspace they span. The points $\mathbf x$ lie on the
  $d$-dimensional real ruled manifold $M$ swept out by the affine $(d-1)$-planes
  $x(t) + M_t$, whose induced metric is allowable; hence $\mathbf x \in \mathcal V_k$.
\end{proof}

In the other direction, Proposition~\ref{prop:conf-envelope} (Proposition 2.3 of
Section~2) shows that the holomorphic envelope of $\mathcal U_k$ contains
$\mathcal V_k^{\mathrm{flat}}$, the part of $\mathcal V_k$ represented by flat affine
submanifolds $M \subset \mathbb M_{\mathbb C}$.

Stating these results in Lean rests on infrastructure not developed here, most of it
deferred already in Section~2. The Wightman permuted extended tube $\mathcal U_k$ and
the notions of holomorphic envelope and domain of holomorphy in
$(\mathbb M_{\mathbb C})^k$ are the tube-domain material that Section~2 deferred;
Proposition~\ref{prop:conf-envelope} is itself not yet formalized. The candidate
$\mathcal V_k$ uses totally real submanifolds of $\mathbb M_{\mathbb C}$ with allowable
induced metric, the configuration space $\mathrm{Conf}_k(\mathbb M_{\mathbb C})$, and
the infinite-dimensional complex manifold $\mathcal F_k$ with its quotient
construction; the map $\mathcal E_k$ rests on the observables $\mathcal O$
(Definition~\ref{def:observables}) and on Principle~5.1. That $\mathcal V_k$ is a
domain of holomorphy remains, in the paper itself, a conjecture.
